Palmer amaranth (*Amaranthus palmeri*) adaptation to agroecossystems

### 2021-08-13 Manuscript/Manuscript.Rmd
as

@@

In the southeastern US, early growing Palmer amaranth is well known to have higher impact on cotton yields than late established plants [@macrae2013]. In the northern states, Palmer amaranth impact on the agroecossystem is recent. Studies investigating Palmer amaranth in those locations is limited due to the plant classification as noxious weed species [@yu2021]. Nonetheless, the continuous Palmer amaranth dispersal and potential establishment across northern United States is concerning and warrant investigations on species morphology in such environments. Understanding Palmer amaranth biology and growing strategies under different agroecossystems can enhance our knowledge on species adaptation. It can also aid on designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact, and design resilient and sustainable farming systems [@maclaren2020]. Therefore, the objective of this study was to investigate the flowering pattern, gender, biomass production, and height of Palmer amaranth growing under corn, soybean and fallow at two timings across five locations in the United States Midwest.

@@

@@

The experiment was conducted in 2018 and 2019 under field conditions at five locations: Arlington (Washington County, Wisconsin), Clay Center (Clay County, Nebraska), Grant (Perkins County, Nebraska), Lincoln (Lancaster County, Nebraska), and Macomb (McDonough County, Illinois).

A glyphosate-resistant soybean cultivar (DSR-1950 R2Y at 296,400 seeds ha 1), and a corn hybrid were planted at

Monthly mean air temperature and sum precipitation were obtained using Daymet weather data from June through September across the five locations in 2018 and 2019 [@correndo2021] (Figure 1)

```{r fig1, echo = FALSE, message=FALSE, fig.show = "hide"}
knitr::include_graphics("../data analysis/weather/Figure 1.png")
```

The field experimental unit were six adjacent 9.1 m wide (12 rows at 72.2 cm row spacing) by 10.7 m long. Each experimental unit was planted with corn or soybean (DSR-1950 R2Y at 296,400 seeds ha 1), or under fallow condition. Palmer amaranth seedlings (potting mix + two seedlings) were and gently transferring to the ground (6 cm deep and 8 cm wide). Twenty-four plants were equidistantly placed (0.76 m apart) between rows within each agroecossystems. After a week, one was eliminated and one was kept. There were two transplant timing: first (June 1) and second (July 1). There were 24 Palmer amaranth plants in each experimental unit, with a total of 144 plants for each location. The study was repeated twice.

@@

Palmer amaranth height and biomass were performed with a linear mixed model using *lmer* function from "lme4" package [@bates2015]. Plant height and biomass were transformed to meet model assumption of normality. In the model, agroecosystem (crop, soybean, fallow) was the fixed effect and year nested with location the random effects. Analysis of variance was performed with *anova* function from "car" package [@fox2018]. Marginal means and compact letter display were estimated with *emmeans* and *cld* from packages "emmeans" and "multcomp" [@hothorn2008].

@@

Palmer amaranth accumulated more biomass when growing in fallow compared to Palmer amaranth growing in soybean and corn (Figure 3A). At first cohort time, Palmer amaranth biomass was 75.5 g plant-1, 28.3 g plant-1 and 16.3 g plant-1 in fallow, soybean and corn, respectively. At second cohort timing, Palmer amaranth produced 62.6 g plant in fallow, followed by 6.3 g plant in soybean, and 1.4 g plant in corn.

Palmer amaranth height was more uniform across cohort timings, except when growing in corn (Figure 3B). Palmer amaranth achieved 69.2 cm tall when growing at first cohort timing in bareground, which was not different to 70.7 cm tall at second cohort timing (P = 0.74). In addition, no difference in Palmer amaranth height (69.3 cm) was found in soybean at first cohort compared to fallow (P > 0.75). At second cohort in soybean, Palmer amaranth was near to 10 cm lower compared to its first cohort time (P = 0.04). The tallest and smallest Palmer amaranth plants were found in corn. Palmer amaranth reached 85.2 cm tall at first cohort and 38.2 cm at second cohort timings.

@@

Palmer amaranth growing in corn resulted in a longer flowering pattern compared to fallow and soybean at first cohort (Figure 2A). Nonetheless, the 10% cumulative Palmer amaranth flowering in soybean, fallow and corn occurred at the end of June. Palmer amaranth reached 10% flowering in soybean at doy 180, which was slightly different from fallow (doy 180.9; *P* = 0.01) and corn (doy 181.7; *P* = 0.00). The 50% Palmer amaranth cumulative flowering occurred in July. For example, Palmer amaranth reached 50% flowering in fallow at doy 193.4, followed by soybean (doy 194.8), corn (doy 206.6). Similar trend was observed at 90% Palmer amaranth cumulative flowering. Palmer amaranth growing in corn reached 90% flowering at doy 252.6 (early September), which was 37.8 and 32.2 days after Palmer amaranth 90% flowering in fallow and soybean, respectively.

```{r fig2, echo = FALSE, message=FALSE, fig.show = "hide"}
knitr::include_graphics("../data analysis/figures/Figure 2.png")
```

Palmer amaranth cumulative flowering at second cohort ranged from mid July to mid September (Figure 2B). Palmer amaranth growing in fallow resulted in earlier flowering time compared to soybean and corn. Palmer amaranth growing in fallow reached 10%, 50%, and 90% flowering time at day 203.8, 214.4, and 232.2, respectively. Palmer amaranth growing in soybean reached 10% flowering at doy 210.9, which was 6 days prior to corn (*P*-value = 0.00). Similar trend was observed at 50% flowering, whereas Palmer amaranth reached 50% flowering in corn (doy 233.0) 4 days after soybeans (doy 228.9; *P* = 0.00). The 90% Palmer amaranth cumulative flowering occurred at same day in corn (260.9) and soybean (260.5; *P* = 0.66).

```{r eval = FALSE, include = FALSE}
options(scipen = 7)
EDcomp(model4, c(90, 90))

result3 %>% 
  filter(ed == "90" & month == "Second transplanting")
```

@@

```{r eval = FALSE, include = FALSE}
multcomp::cld(analysis$emmeans[[2]], 
              alpha=0.05, Letters=letters, 
              adjust="none", reversed = TRUE)
```

```{r fig3, echo = FALSE, message=FALSE, fig.show = "hide"}
knitr::include_graphics("../data analysis/figures/Figure 3.png")
```

```{r fig4, echo = FALSE, message=FALSE, fig.show = "hide"}
knitr::include_graphics("../data analysis/figures/image.jpeg")
```

@@

An ecological approach to reduce seed production in Palmer amaranth is understanding its flowering pattern. Our study suggests that Palmer amaranth flowering pattern was slightly influenced by cropping systems and cohort timings. Palmer amaranth growing in fallow and corn resulted in the overall shortest and longest flowering pattern, respectively. When growing in soybean, Palmer amaranth flowering pattern was similar to bareground at first cohort but similar to corn at second cohort timing. Plant flowering initiation is complex and it depends on biological and ecological factors [@lang1965]. We hypothesize that when growing in high competition (e.g., second cohort), Palmer amaranth plants tend to initiate flowering early, as well as having an extended flowering pattern. Early flower initiation is plant strategy when growing in stress conditions. For example, when growing under water stress, early flowering in Palmer amaranth resulted in a mismatch between female and male plants by seven days [@mesgaran2021]. A mismatch in Palmer amaranth male and female flowering period can minimize plant outcross, and thus reduce plant seed production. Sex dimorphism manipulation is considered a potential ecological pest control [@mcfarlane2018; @schliekelman2005].

@@

Our study demonstrated the short-term Palmer amaranth plasticity to grow and develop into cropping-systems. Palmer amaranth management priority should focus on minimizing Palmer amaranth dispersal into new geographies. Flower manipulation is a promising ecological Palmer amaranth but it is not well understood. Nonetheless, early-season management programs would have a large negative effect in Palmer amaranth growth and development. Tactics that promote early-season crop advantage against Palmer amaranth, including early crop planting, crop rotation [@oliveira2021a], plant width, preemergence applied herbicide [@sanctis2021], and crop residue (e.g. cover crops) would minimize the negative impact of Palmer amaranth in agroecossystems.

# Disclosure/Conflict-of-Interest Statement {-}

<!--  
Frontiers follows the recommendations by the International Committee of Medical
Journal Editors (http://www.icmje.org/ethical_4conflicts.html) which require
that all financial, commercial or other relationships that might be perceived by
the academic community as representing a potential conflict of interest must be
disclosed. If no such relationship exists, authors will be asked to declare that
the research was conducted in the absence of any commercial or financial
relationships that could be construed as a potential conflict of interest. When
disclosing the potential conflict of interest, the authors need to address the
following points:

@@

```{r, Figure-2, ref.label = "fig2", echo = FALSE, message = FALSE, fig.height=9, fig.width=4, fig.align='center', fig.cap='Cumulative flowering of Palmer amaranth at first and second transplant timing (A) and day of year of 10, 50, and 90 cumulative flowering at first and second transplant timing (B)', out.width = "130mm", out.height = "130mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```

```{r, Figure-3, ref.label = "fig3", echo = FALSE, message = FALSE, fig.height=9, fig.width=4, fig.align='center', fig.cap='Palmer amaranth biomass (A) and height (B) growing in corn, fallow, and soybean across Arlington, Clay Center, Grant, Lincoln, and Macomb', out.width = "130mm", out.height = "70mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```

```{r, Figure-4, ref.label = "fig4", echo = FALSE, message = FALSE, fig.height=9, fig.width=4, fig.align='center', fig.cap='Harvest Palmer amaranth plants at 40 days after first transplant timing. From left to right, Palmer amaranth growing in fallow, soybean and corn in Arlington, Wisconsin', out.width = "70mm", out.height = "70mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```

>>>>

```{r include=FALSE}
library(kableExtra)
library(tidyverse)
```

@@

In the southeastern US, early growing Palmer amaranth is well known to have higher impact on cotton yields than late established plants [@macrae2013]. In the northern states, Palmer amaranth impact on the agroecossystem is recent. Studies investigating Palmer amaranth in those locations is limited due to the plant classification as noxious weed species [@yu2021]. Nonetheless, the continuous Palmer amaranth dispersal and potential establishment across northern United States is concerning and warrant investigations on species morphology in such environments. Understanding Palmer amaranth biology and growing strategies under different agroecossystems can enhance our knowledge on species adaptation. It can also aid on designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact, and design resilient and sustainable farming systems [@maclaren2020]. Therefore, the objective of this study was to investigate the flowering pattern, gender, biomass production, and height of Palmer amaranth growing under corn, soybean and bareground at two timings across five locations in the United States Midwest.

@@

\vspace{0.4cm}

@@

The experiment was conducted in 2018 and 2019 under field conditions at five locations: Arlington, WI (43°18'N, 89°29'W), Clay Center, NE ('N, 'W), Grant (('N, 'W)), Lincoln (('N, 'W)), and Macomb (('N, 'W)).

Fields were conventionally tilled and disked prior to planting. Corn hybrid and soybean varieties were planted in 76-cm row spacing (Table 1). Monthly mean air temperature and sum precipitation were obtained using Daymet weather data from June through September across the five locations in 2018 and 2019 [@correndo2021] (Figure 1)

```{r echo=FALSE, warning=FALSE, message=FALSE}
options(knitr.kable.NA = '')
read_csv("../data/table.csv") %>% 
  janitor::row_to_names(row_number = 1) %>% 
  kable(booktabs = TRUE, caption = "Field study attributes ",
        col.names = c("Attributes",
                           "",
                           "Arlington, WI",
                           "Clay Center, NE",
                           "Grant, NE",
                      "Lincoln, NE", "Macomb, IL")) %>% 
#  pack_rows(
 #   index = c("Bareground" = 2, "Corn" = 4, "Soybean" = 8)) %>% 
  kable_styling(font_size = 12, 
                full_width = FALSE, 
                position = "center",
                latex_options = c("scale_down", "striped", "hold_position")) %>% 
  collapse_rows(columns = 1, latex_hline = "none", valign = "middle") 
```

The field experimental unit were six adjacent 9.1 m wide (12 rows at 72.2 cm row spacing) by 10.7 m long. Each experimental unit was planted with corn or soybean, or under bareground condition. Palmer amaranth seedlings (potting mix + two seedlings) were and gently transferring to the ground (6 cm deep and 8 cm wide). Twenty-four plants were equidistantly placed (0.76 m apart) between rows within each agroecossystems. After a week, one was eliminated and one was kept. There were two transplant timing: first (June 1) and second (July 1). There were 24 Palmer amaranth plants in each experimental unit, with a total of 144 plants for each location. The study was repeated twice.

```{r table1, echo = FALSE, message=FALSE, fig.show = "hide"}
knitr::include_graphics("../data analysis/weather/Figure 1.png")
```

@@

Palmer amaranth height and biomass were performed with a linear mixed model using *lmer* function from "lme4" package [@bates2015]. Plant height and biomass were transformed to meet model assumption of normality. In the model, agroecosystem (crop, soybean, bareground) was the fixed effect and year nested with location the random effects. Analysis of variance was performed with *anova* function from "car" package [@fox2018]. Marginal means and compact letter display were estimated with *emmeans* and *cld* from packages "emmeans" and "multcomp" [@hothorn2008].

@@

Palmer amaranth accumulated more biomass when growing in bareground compared to Palmer amaranth growing in soybean and corn (figure 2A). At first cohort time, Palmer amaranth biomass was 75.5 g plant-1, 28.3 g plant-1 and 16.3 g plant-1 in bareground, soybean and corn, respectively. At second cohort timing, Palmer amaranth produced 62.6 g plant in bareground, followed by 6.3 g plant in soybean, and 1.4 g plant in corn.

Palmer amaranth height was more uniform across cohort timings, except when growing in corn (figure 2B). Palmer amaranth achieved 69.2 cm tall when growing at first cohort timing in bareground, which was not different to 70.7 cm tall at second cohort timing (P = 0.74). In addition, no difference in Palmer amaranth height (69.3 cm) was found in soybean at first cohort compared to bareground (P > 0.75). At second cohort in soybean, Palmer amaranth was near to 10 cm lower compared to its first cohort time (P = 0.04). The tallest and smallest Palmer amaranth plants were found in corn. Palmer amaranth reached 85.2 cm tall at first cohort and 38.2 cm at second cohort timings.

\vspace{0.4cm}

@@

Palmer amaranth growing in corn resulted in a longer flowering pattern compared to bareground and soybean at first cohort (figure3A). Nonetheless, the 10% cumulative Palmer amaranth flowering in soybean, bareground and corn occurred at the end of June. Palmer amaranth reached 10% flowering in soybean at doy 180, which was slightly different from bareground (doy 180.9; *P* = 0.01) and corn (doy 181.7; *P* = 0.00). The 50% Palmer amaranth cumulative flowering occurred in July. For example, Palmer amaranth reached 50% flowering in bareground at doy 193.4, followed by soybean (doy 194.8), corn (doy 206.6). Similar trend was observed at 90% Palmer amaranth cumulative flowering. Palmer amaranth growing in corn reached 90% flowering at doy 252.6 (early September), which was 37.8 and 32.2 days after Palmer amaranth 90% flowering in bareground and soybean, respectively.

Palmer amaranth cumulative flowering at second cohort ranged from mid July to mid September (figure3B). Palmer amaranth growing in bareground resulted in earlier flowering time compared to soybean and corn. Palmer amaranth growing in bareground reached 10%, 50%, and 90% flowering time at day 203.8, 214.4, and 232.2, respectively. Palmer amaranth growing in soybean reached 10% flowering at doy 210.9, which was 6 days prior to corn (*P*-value = 0.00). Similar trend was observed at 50% flowering, whereas Palmer amaranth reached 50% flowering in corn (doy 233.0) 4 days after soybeans (doy 228.9; *P* = 0.00). The 90% Palmer amaranth cumulative flowering occurred at same day in corn (260.9) and soybean (260.5; *P* = 0.66).

\vspace{0.4cm}

@@

```{r fig2, echo = FALSE, message=FALSE, fig.show = "hide"}
knitr::include_graphics("../data analysis/figures/Figure 3.png")
```

```{r fig3, echo = FALSE, message=FALSE, fig.show = "hide"}
knitr::include_graphics("../data analysis/figures/image.jpeg")
```

```{r fig4, echo = FALSE, message=FALSE, fig.show = "hide"}
knitr::include_graphics("../data analysis/figures/Figure 2.png")
```

@@

An ecological approach to reduce seed production in Palmer amaranth is understanding its flowering pattern. Our study suggests that Palmer amaranth flowering pattern was slightly influenced by cropping systems and cohort timings. Palmer amaranth growing in bareground and corn resulted in the overall shortest and longest flowering pattern, respectively. When growing in soybean, Palmer amaranth flowering pattern was similar to bareground at first cohort but similar to corn at second cohort timing. Plant flowering initiation is complex and it depends on biological and ecological factors [@lang1965]. We hypothesize that when growing in high competition (e.g., second cohort), Palmer amaranth plants tend to initiate flowering early, as well as having an extended flowering pattern. Early flower initiation is plant strategy when growing in stress conditions. For example, when growing under water stress, early flowering in Palmer amaranth resulted in a mismatch between female and male plants by seven days [@mesgaran2021]. A mismatch in Palmer amaranth male and female flowering period can minimize plant outcross, and thus reduce plant seed production. Sex dimorphism manipulation is considered a potential ecological pest control [@mcfarlane2018; @schliekelman2005].

@@

Our study demonstrated the short-term Palmer amaranth plasticity to grow and develop into cropping-systems. Is likely that Palmer amaranth range will continue to expanding to new geographies. Therefore, preventive management is a priority to minimizing Palmer amaranth dispersal. Reactive management should focus on early-season management programs, which would have a large negative effect on Palmer amaranth growth and development. Tactics that promote early-season crop advantage against Palmer amaranth, including early crop planting, crop rotation [@oliveira2021a], plant width, preemergence applied herbicide [@sanctis2021], and crop residue (e.g. cover crops) would minimize the negative impact of Palmer amaranth in agroecossystems. The aggressiveness and differential Palmer amaranth adaptation to agroecosystem is striking and require national efforts to minimize the species impact on economy and sustainability.

# Disclosure/Conflict-of-Interest Statement {-}  
Flower manipulation is a promising ecological Palmer amaranth but it is not well understood.
<!--  
Frontiers follows the recommendations by the International Committee of Medical
Journal Editors (http://www.icmje.org/ethical_4conflicts.html) which require
that all financial, commercial or other relationships that might be perceived by
the academic community as representing a potential conflict of interest must be
disclosed. If no such relationship exists, authors will be asked to declare that
the research was conducted in the absence of any commercial or financial
relationships that could be construed as a potential conflict of interest. When
disclosing the potential conflict of interest, the authors need to address the
following points:

@@

```{r, Figure-2, ref.label = "fig2", echo = FALSE, message = FALSE, fig.height=9, fig.width=4, fig.align='center', fig.cap='Palmer amaranth biomass (A) and height (B) growing in corn, bareground, and soybean across Arlington, WI, Clay Center, NE, Grant, NE, Lincoln, NE and Macomb, IL', out.width = "150mm", out.height = "90mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```

```{r, Figure-3, ref.label = "fig3", echo = FALSE, message = FALSE, fig.height=9, fig.width=4, fig.align='center', fig.cap='Harvest Palmer amaranth plants at 40 days after first transplant timing. From left to right, Palmer amaranth growing in bareground, soybean and corn in Arlington, Wisconsin', out.width = "90mm", out.height = "90mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```

```{r, Figure-4, ref.label = "fig4", echo = FALSE, message = FALSE, fig.height=9, fig.width=4, fig.align='center', fig.cap='Cumulative flowering of Palmer amaranth at first and second transplant timing (A) and day of year of 10, 50, and 90 cumulative flowering at first and second transplant timing (B)', out.width = "150mm", out.height = "150mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```
